---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r warning = FALSE}
dataFile <- unz(description = "activity.zip", filename = "activity.csv")
activityData <- read.table(file = dataFile, sep = ",", header = TRUE, na.strings = "NA")
activityData$date <- as.Date(activityData$date, format = "%Y-%m-%d")
## melt data for later use
library(reshape2)
actMelt <- melt(activityData, id = c("date", "interval"), measure.vars = c("steps"))
```

## What is mean total number of steps taken per day?

```{r}
dateSums <- dcast(actMelt, date ~ variable, sum, na.rm = TRUE)
plot(steps ~ date, data = dateSums, type = "h", ylab = "Steps", xlab = "Date")
mean(xtabs(steps ~ date, data = activityData))
mean(dateSums$steps, na.rm = TRUE)
median(xtabs(steps ~ date, data = activityData))
median(xtabs(steps ~ date, data = activityData), na.rm = TRUE)
median(dateSums$steps, na.rm = TRUE)
```

## What is the average daily activity pattern?
```{r}
intervalMeans <- dcast(actMelt, interval ~ variable, mean, na.rm = TRUE)
plot(steps ~ interval, data = intervalMeans, type = "l")
max(intervalMeans$steps)
intervalMeans[intervalMeans$steps == max(intervalMeans$steps),]
```


## Imputing missing values
```{r}
sum(is.na(activityData$steps))

newCol <- rep(floor(intervalMeans$steps), 61)
imputedData <- cbind(activityData, newCol)
imputedData$newCol[which(!is.na(imputedData$steps))] <- imputedData$steps[which(!is.na(imputedData$steps))]
library(reshape2)
impMelt <- melt(imputedData, id = c("date", "interval"), measure.vars = c("newCol"))
dateSums <- dcast(impMelt, date ~ variable, sum, na.rm = TRUE)
plot(newCol ~ date, data = dateSums, type = "h", ylab = "Steps", xlab = "Date")
mean(xtabs(newCol ~ date, data = imputedData))
mean(dateSums$newCol, na.rm = TRUE)
median(xtabs(newCol ~ date, data = imputedData))
median(xtabs(newCol ~ date, data = imputedData), na.rm = TRUE)
median(dateSums$newCol, na.rm = TRUE)

``` 
Using the floor function as step data should be in whole numbers

## Are there differences in activity patterns between weekdays and weekends?
```{r warning = FALSE, message = FALSE}
library(dplyr)
activityData <- mutate(activityData, days = factor(1*(weekdays(date) %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")), labels = c("weekday", "weekend")))
actMelt <- melt(activityData, id = c("date", "interval", "days"), measure.vars = c("steps"))
intervalMeans <- dcast(actMelt, interval + days ~ variable, mean, na.rm = TRUE)
par(mfrow = c(2,1))
plot(steps ~ interval, data = intervalMeans[intervalMeans$days == "weekend",], type = "l")
plot(steps ~ interval, data = intervalMeans[intervalMeans$days == "weekday",], type = "l")
```

## exploratory scripting

head(activityData)
str(activityData)
xtabs(steps ~ date, data = activityData)
plot(xtabs(steps ~ date, data = activityData), type = "h", ylab = "Steps", xlab = "Date")
intervalMeans[activityData[is.na(activityData$steps),3]==intervalMeans$interval,2]
17568/288
[1] 61
newCol <- rep(floor(intervalMeans$steps), 61)
imputedData <- cbind(activityData, newCol)
